\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-11-01 demystify.tex
updates
Updates from Overleaf

Choosing a school and an advisor is clearly one of the most important things in your mind when you apply to and especially when you get admitted.  This is further complicated due to cultural differences and unfamiliarity of  international students to the education system in the .  This section aims to mitigate some confusions and help you make informed decision.

@@

This section aims to explain different types of universities in the US in the context of obtaining a CS Ph.D degree.

\subsection{Schools that offer Ph.D. in CS}

Most universities the US will have CS programs. 
However, while many universities offer Ph.D. in their CS programs, many do not.  These universities might offer just Bachelor degrees (e.g., BS) and no graduate studies (i.e., no MS or Ph.D degrees), or they just offer MS programs (but no PhD). For example, Penn State in University Park has Ph.D. in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiplication locations of the University of Texas, e.g., Austin, Dallas, Arlington, all have PhDs in CS.

@@

\subsection{Selecting and Ranking Schools for CS}
   \begin{center}
    \includegraphics[scale=0.5]{c1.png}
   \end{center}

@@

\subsubsection{Tenured, Tenure Track, Teaching,  Assistant Profs., Associate Profs}
There are several types of faculty in a university that has a CS PhD program.  You might hear about tenured,  tenure track, teaching faculty.  You might also see assistant, associate, full, adjunct, emeritus, teaching, research professors. Here is a primer on these titles.

@@

\paragraph{Rankings} Faculty have a ranking, e.g., assistant, associate, full, regardless if they are tenure line or teaching.  Briefly, assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate).

@@

%The tenure system in universities in the US (and other countries) can be confusing. In most cases, you probably don't need to know much about it. But it is helpful to know a few things so you know how to find PhD advisor. First, faculty of a CS (and other) department can be divided into 2 types:  (i) tenured and tenure-track and (ii) non-tenured or teaching faculty.  Tenured/Tenure -track faculty focus on research and related tasks such as mentoring and getting grants to support students, etc. Their teaching load is very light, typically they just teach 1 or at most 2 classes a semester.  Non-tenured/teaching faculty focuses on teaching, and often teach 3-4 classes per semester.  They can mentor Ph.D. students but they typically cannot serve as a formal advisor of Ph.D. students (i.e., they cannot graduate PhD students). Note that there is also affiliated faculty who is tenured/tenure-track profs. from another department. They can serve in Ph.D. dissertation committee but they likely cannot serve as the main advisor of CS students.

@@

\subsection{Should you contact a professor? What to do to get a desired reply?}\label{sec:contact}

@@

@@

\subsection{PhD in other related fields: CE, IST, Cybersecurity}

@@

Furthermore, avoid using Mr. or Mrs., or the professor's first name if you're not acquainted with them yet.  As you become more familiar with a professor and depending on their preferences, you may transition to addressing them by their first name.
For example, my students and colleague call me Vu. Some students call me \emph{Dr. Vu}, which I find a bit amusing but am totally fine with it.

@@

\paragraph{Referring to professors you know} Because you are already familiar with these individuals, you can just informally use their firstname as mentioned above (or Dr./Prof., if you want to be formal). You can also include their institution if it makes it more precise.  For example, I did my postdoc with Jeff Foster at Univ. of Maryland.   Do not include ranking (e.g., Assistant, Associate, Scientist, ...) when referring to someone. I see many international students include a lengthy title of people they know, e.g., I am advised by Asst. Prof. X, and I also collaborate with Distinguished Scientist Y.  This is \emph{not necessary} and makes it look like you're trying to show off your connections. In short, these nuances represent some cultural differences that you may encounter and will gradually adapt to. More on cultural differences in \S\ref{sec:cultural}.

>>>>

Choosing a school and an advisor is clearly one of the most important things in your mind when you apply to and especially when you get admitted.  This is further complicated due to cultural differences and unfamiliarity of  international students to the US higher education system.  This section aims to mitigate some confusions and help you make informed decision.

@@

We will first discuss about universities in the US that offer PhD in CS.

\subsubsection{Schools that offer PhD in CS}

Most universities the US will have CS programs. 
However, while many universities offer PhD in their CS programs, many do not.  These universities might offer just Bachelor degrees (e.g., BS) and no graduate studies (i.e., no MS or Ph.D degrees), or they just offer MS programs (but no PhD). For example, Penn State in University Park has Ph.D. in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiplication locations of the University of Texas, e.g., Austin, Dallas, Arlington, all have PhDs in CS.

@@

\subsubsection{Selecting and Ranking Schools for CS}
   \begin{center}
    \includegraphics[scale=0.5]{c1.png}
   \end{center}

@@

Now we discuss about finding advisors.

\subsubsection{Tenured, Tenure Track, Teaching,  Assistant Profs., Associate Profs}
There are several types of faculty in a university that has a CS PhD program.  You might hear about tenured,  tenure track, teaching faculty.  You might also things such as assistant, associate, full, adjunct, emeritus, teaching, research professors. Here is a primer on these terminologies.

@@

\paragraph{Rankings} Faculty have a ranking, e.g., assistant, associate, full, regardless if they are tenure line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate).

@@

@@

\subsubsection{Should you contact a professor? What to do to get a desired reply?}\label{sec:contact}

@@

\subsubsection{PhD in other related fields: CE, IST, Cybersecurity}

@@

@@

Furthermore, avoid using Mr. or Mrs., or the professor's first name if you're not acquainted with them yet.  As you become more familiar with your prof and depending on their preferences, you may transition to addressing them by their first name.
For example, I prefer that my students and colleague call me Vu. Some students call me \emph{Dr. Vu}, which I find a bit amusing but am totally fine with it.

@@

\paragraph{Referring to professors you know} Because you are already familiar with these individuals, you can just informally use their firstname as mentioned above (or Dr./Prof., if you want to be formal). You can also include their institution if it makes it more precise.  For example, I can say:  \emph{"I did my postdoc with Jeff Foster at Univ. of Maryland"}.

Do not include ranking (e.g., Assistant, Associate, Scientist, ...) when referring to someone. I see many international students include a lengthy title of people they know, e.g., \emph{I am advised by Asst. Prof. X, and I also collaborate with Distinguished Scientist Y}.  This is \emph{not necessary} and makes it look like you're trying to show off your connections. These nuances represent some cultural differences that you may encounter and will gradually adapt to. More on cultural differences in \S\ref{sec:cultural}.
